---
title: "Archaeological Data"
author: Sydney Mylan & Corina Amundson
format: dashboard
---

```{r setup}
#| warning: false
#| message: false
#| output: false
library(ggraph)
library(igraph)
library(tidyverse)
library(RColorBrewer)
library(readxl)
library(usmap)
library(ggplot2)
library(GGally)
library(maps)
library(readr)
library(knitr)
library(stringr)
library(forcats)
library(easyalluvial)

#Read in Data
arch_data <- read_csv("new.csv")

#Cleaning the data

#Renaming Columns
names(arch_data) = c("Subject", "Time", "Type", "Class", 
                     "Description", "Material", "Name", "Quantity", 
                     "Collector", "Locality", "Country", "State", 
                     "City", "SiteID", "SiteName", "Map", "Remarks")

#Assigning Subject as numerical for data cleaning
arch_data_fixed <- arch_data |>
  mutate(Subject = as.numeric(Subject))

#Removing Unnecessary Variables
arch_data_fixed <-  subset(arch_data_fixed, select = 
                             -c(SiteID, SiteName, Map, Remarks, Locality, Collector))


#Data cleaning for Geographical Mapping

#mutating states to match each other
arch_data_fixed = arch_data_fixed |>
  mutate(State = case_when(
    State %in% c("NC") ~ "North Carolina",
    State == "SC" ~ "South Carolina",
    State == "OH" ~ "Ohio",
    State == "FL" ~ "Florida",
    State == "VA" ~ "Virginia",
    State == "Virginia/NC" ~ "Virginia",
    .default = State
  ))

#Data visualization 
Arch_Table <- head(arch_data_fixed, 10)

#filtering by state and quantity into a new data set
map_data <- arch_data_fixed |> 
  select(State | Quantity)  

#ordering alphabetically
map_data <- map_data[order(map_data$State), ]

#labeling non states as NA and removing them
map_data = map_data |>
  mutate(State = case_when(
    State == "Zaire" ~ NA,
    State == "Finchaven" ~ NA,
    .default = State
  ))
map_data <- na.omit(map_data)

#Summarizing quantity by state
map_data <- map_data |>
  group_by(State) |>
  summarise(S_Quantity = sum(Quantity, na.rm = TRUE))

#Adding in States as NA that were not included in dataset
map_data_fixed <- map_data |> add_row(
  State = "Alabama", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "Alaska", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "Colorado", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "District of Columbia", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "Hawaii", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "Idaho", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "Louisiana", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "Maine", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "Massachusetts", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "Montana", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "Nebraska", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "New Hampshire", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "North Dakota", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "Rhode Island", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "South Dakota", S_Quantity = NA)

map_data_fixed <- map_data_fixed |> add_row(
  State = "Wyoming", S_Quantity = NA)

#Ordering all States Alphabetically
map_data_fixed <- map_data_fixed[order(map_data_fixed$State), ]

#add in Puerto Rico after the alphabetization to match statepop
map_data_fixed <- map_data_fixed |> add_row(
  State = "Puerto Rico", S_Quantity = NA)
#Plotting

#matching arch data to statepop
map_data_fixed$fips <- statepop$fips
map_data_fixed$abbr <- statepop$abbr

#Plot all States where Artifacts were Observed
USA_MAP <- plot_usmap(data = map_data_fixed,
           values = "S_Quantity") +
  labs(title = "States where Artifacts were Observed") + 
  theme(panel.background = element_rect(colour = "black"))+
  scale_fill_continuous(low = "lightblue", high ="lightblue", 
                        name = "Quantity",label = scales::comma)+
  theme(legend.position = "FALSE")

#Plot Bar Plot with top 5 States by Quantity
map_data2 <- map_data |> 
  filter(S_Quantity > 500) 
map_data2 <- map_data2[order(map_data2$S_Quantity), ]

BAR_PLOT <- ggplot(map_data2, aes(x=State, y=S_Quantity)) + 
  geom_bar(fill = "tomato", stat = "identity") + labs(
    x = "Number of Artifacts Observed",
    y = "State",
    title = "Top 5 States with the most Artifacts") +
  geom_text(aes(label = S_Quantity), vjust = -0.2, size = 8) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


#Plot Gradient Map by States and Quantity (0-100)
USA_0100_MAP <- plot_usmap(data = map_data_fixed,
                    values = "S_Quantity") +
  labs(title = "States Where Less Than 100 Artifacts were Observed") + 
  theme(panel.background = element_rect(colour = "black"))+
  scale_fill_continuous(low = "lightblue", high ="tomato", 
                        name = "Quantity",label = scales::comma,
                        limits = c(0,100)) + 
  theme(legend.position = "right")


#More data cleaning for Corina's plots 

arch_data_clean <- arch_data_fixed %>%
  mutate(Material_clean = Material %>%
           str_to_lower() %>%
           str_trim() %>%
           str_replace_all("[[:punct:]]+", " ") %>%
           str_squish())

arch_data_clean <- arch_data_clean %>%
  filter(
    !is.na(Type), Type != "",
    !is.na(Class), Class != "",
    !is.na(Material_clean), Material_clean != ""
  )

#Grouping variables for dendrogram
arch_clean_1 <- arch_data_clean |>
  mutate(Type = fct_collapse(Type, UNIDENTIFIED = c("UNIDENTIFIED MATERIAL", "COMPOSITE"), HUMAN = "HUMAN REMAINS"))

arch_clean_2 <- arch_clean_1 |>
  mutate(Class = fct_collapse(Class, BONE = c("BONE", "OSTEOLOGICAL", "ANTLER"), CLAY = c("CLAY", "CERAMIC"), OTHER_MINERAL = "OTHER MINERAL MATERIALS", OTHER_ANIMAL = "OTHER ANIMAL MATERIALS", OTHER_PLANT = "OTHER PLANT MATERIALS"))

arch_clean_3 <- arch_clean_2 |>
  mutate(Material_clean = fct_collapse(Material_clean, Lithic = c("mixed lithic", "lithic", "unidentified lithic", "unspecified lithic"), Quartz = c("crystal quartz", "quartz", "quartzite"), Chert_Flint = c("chert", "flint english", "cryptocrystalline"), Chalcedony_Jasper = c("chalcedony", "jasper"), Volcanic = c("rhyolite", "metarhyolite", "basalt"), Coarse_Ingeous = c("granite", "gabbroid"), Sedimentary = c("sandstone", "shale"), Metamorphic = c("slate", "argillite"), Soft_Stone = c("steatite", "soapstone", "catlinite"), Mineral_Pigments = c("ochre", "hematite", "limonite"), Other_Minerals = c("mica", "greenstone", "ironstone"), Ceramic = c("ceramic", "ceramic porcelain", "ceramic coarse ew", "ceramic refined ew"), Architectural_Ceramic = c("brick", "clay", "kaolin"), Unidentified = c("ceramic unidentifie", "unknown material", "ceramic unidentifi", "unspecified plant", "unidentified", "undetermined", "miscellaneous"), Iron = c("iron", "iron and brick"), Non_Ferrous_Metal = c("copper alloy", "copper", "lead", "aluminum"), Industrial_Byproducts = c("slag", "clinker"), Metal = c("metal", "metal unidentified", "cast"), Glass = c("milk glass", "glass"), shell = c("shell", "shell and bone"), Teeth = c("ivory", "tooth", "teeth"), Organic = c("coal", "charcoal", "organic"), Botanical = c("wood", "bark", "seed", "nut", "vegetation"), Synthetic = c("paper", "plastic"), Construction = c("mortar", "plaster"), Mixed_Materials = c("quartz mixed lithic", "chert mixed lithic", "mixed material", "multiple materials")))

#narrowing data to just Virginia
virginia_data <- arch_data_clean |>
  filter(State == "Virginia")

#narrowing data to just Virginia where the type is always Bones
bone_virginia_data <- virginia_data |>
  filter(Type == "ANIMAL")

#narowing data do just bones in Virginia where both material and type is bones 
bones_virginia_data <- bone_virginia_data |>
  filter(Material_clean == "bone")

#Filtering Artifacts with turt in the name
Turtle <- bones_virginia_data %>%
  filter(str_detect(Name, regex("turt", ignore_case = TRUE)))

#Filtering artifact with owl in the name
owl_rows <- bones_virginia_data %>%
  filter(str_detect(Name, regex("owl", ignore_case = TRUE)))

#Combining missppelling and getting rid of eastern meadowlark to just have owls
owl_rows_1 <- owl_rows |>
  mutate(Name = fct_collapse(
    Name,
    `eastern screech owl`= c("Eastern screech owl", "eastern screech owl"),
    Barred_owl = "Barred owl"
  )) |>
  filter(Name != "Eastern meadowlark")
```


# Overview


The Archaeological Scientific Collection from the Virginia Museum of Natural History contains more than 450,000 cataloged artifacts and samples documenting human history, material culture, and archaeological research across different geographic regions and time periods. The dataset is organized through multiple quantitative and categorical variables.



```{r}
#| title: Overview of Data
kable(Arch_Table)
```


# Geographic

## Row {width=60%}
```{r}
USA_MAP
```

```{r fig.height = 10, fig.width = 12}
BAR_PLOT
```
## Row {width=40%}

```{r}
USA_0100_MAP
```

::: {.card}

:::

# Artifacts

## Row {height="50%"}

```{r}
#| title: Alluvial Diagram
#alluvial plot with just basic cleaning
arch_data_clean %>% dplyr::select(Type, Class, Material_clean) %>% alluvial_wide()
```

```{r}
#| title: Type -> Class -> Material
## Deprogrammed for cleaned and grouped materials 
d1 <- arch_clean_3 |>
  distinct(Type) |>
  transmute(from = "origin", to = Type)

d2 <- arch_clean_3 |>
  distinct(Type, Class) |>
  transmute(from = Type, to = Class)

d3 <- arch_clean_3 |>
  distinct(Class, Material_clean) |>
  transmute(from = Class, to = Material_clean)

edges <- bind_rows(d1, d2, d3) |>
  filter(!is.na(from), !is.na(to))

mygraph <- graph_from_data_frame(edges)

ggraph(mygraph, layout = 'dendrogram', circular = TRUE) + 
  geom_edge_diagonal(colour="grey") +
  scale_edge_colour_distiller(palette = "RdPu") +
   geom_node_text(aes(x = x*1.15, y=y*1.15, label = name, angle = node_angle(x, y)), size = 2.7, ) +
  geom_node_point(aes(filter = leaf, x = x*1.07, y=y*1.07, colour="d1", size= 2, alpha=0.2)) +
  scale_colour_manual(values= rep( brewer.pal(9,"Paired") , 30)) +
  scale_size_continuous( range = c(0.1,10) ) +
  theme_void() +
  theme(
    legend.position="none",
    plot.margin=unit(c(0,0,0,0),"cm"),
  ) +
  expand_limits(x = c(-1.3, 1.3), y = c(-1.3, 1.3))
```

# Virginia Bones

## Row {height="50%"}
```{r}
#| title: Virginia
#ungrouped dendragram of just Virginia
d4 <- virginia_data |>
  distinct(Type) |>
  transmute(from = "origin", to = Type)

d5 <- virginia_data |>
  distinct(Type, Class) |>
  transmute(from = Type, to = Class)

d6 <- virginia_data |>
  distinct(Class, Material_clean) |>
  transmute(from = Class, to = Material_clean)

edges1 <- bind_rows(d4, d5, d6) |>
  filter(!is.na(from), !is.na(to))

mygraph1 <- graph_from_data_frame(edges1)

ggraph(mygraph1, layout = 'dendrogram', circular = TRUE) + 
  geom_edge_diagonal(colour="grey") +
  geom_node_point(aes(color = name), size = 2, show.legend = FALSE) +
  geom_node_text(aes(label = name, angle = node_angle(x, y)), size = 2.5, hjust = 0.5) +
  theme_void()
```

```{r}
#| title: Bones in Virginia
# Dendrogram of the animals in Virginia
d8 <- bone_virginia_data |>
  distinct(Class) |>
  transmute(from = "origin", to = Class)

d9 <- bone_virginia_data |>
  distinct(Class, Material_clean) |>
  transmute(from = Class, to = Material_clean)

edges2 <- bind_rows(d8, d9) |>
  filter(!is.na(from), !is.na(to))

mygraph2 <- graph_from_data_frame(edges2)

ggraph(mygraph2, layout = 'dendrogram', circular = TRUE) + 
  geom_edge_diagonal(colour="grey") +
  geom_node_point(aes(color = name), size = 2, show.legend = FALSE) +
  geom_node_text(aes(label = name, angle = node_angle(x, y)), size = 3.5, hjust = 0.5) +
  theme_void()
```

## Row {height="50%"}

```{r}
#| title: Turtle
#Bar Chart of Bones in Virginia that are turtles 
ggplot(Turtle, aes(x = Name)) +
  geom_bar(aes(fill = Name)) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "none")
```

```{r}
#| title: Owls
#Bar Chart of Bones in Virginia that are owls 
ggplot(owl_rows_1, aes(x = Name)) +
  geom_bar(aes(fill = Name)) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "none")
```

# Conclusion

## Row {height="80%"}

::: {.card title="Conclusion"}
Write conclusion
:::

::: {.card title="Work Cited"}
Website to Virginia Museum of Natural History:
https://www.vmnh.net/research-collections/scientific-collections 

Sydney's & Corina's Repo:
https://github.com/SydneyMylan/408-Project-Arch-Data

Deprogram Resource:
https://r-graph-gallery.com/dendrogram.html

R Color Table:
https://cocolaberica.github.io/rcolortable 
:::

## Row {height="20%"}
::: {.card}
Thank You!
:::
